Summary of CA Rice Breeding

### 2017-04-25 output/breed_paper.Rnw
Draft of paper before removing report details

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{Sweave}
\linespread{1.3}

@@

\section{Background}

In general, rice yields in CA have been flat over the last 20+ years,
compared to the preceding period. Yet, annual reports from the CA
Cooperative Rice Research Foundation show that each released cultivar
has a yield advantage over the previous cultivars, typically on the
order of 4--6\%. This raises the question of how to reconcile these
facts. Namely, there are two possibilities:

@@

plot(yield_lb ~ yrs_in_trial, data = mod_data, pch = 16, col = rgb(0,0,0,0.25),
     xlab = "Year since entry",
     ylab = expression("Yield ( lbs acre"^-1~")"))
lines(x = yrs_in, y = ints[2,], lty = 2) 
polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
text(20, 3500, bquote(.(round(fixef(yrs_fit)[2] * 1000, 1))~lbs~year^-1))

@ 
\caption{There is a small decrease in yield over the time since a
  cultivar was entered into the trial.}
\label{fig:yrs}
\end{figure}

As seen in Fig. \ref{fig:yrs}, there is a small decrease in yield
over time once a cultivar is released. This decrease is estimated to be
\Sexpr{round(fixef(yrs_fit)[2] * 1000, 1)} lbs
year\textsuperscript{-1}. The oldest cultivar in the trials has lost
an estimated \Sexpr{round(fixef(yrs_fit)[2] * 1000 * 36, 1)} lbs
acre\textsuperscript{-1} since release.

@@

plot(yield_lb ~ release_yr, data = mod_data, pch = 16, col = rgb(0,0,0,0.25),
     xlab = "Year of Release",
     ylab =  expression("Yield ( lbs acre"^-1~")"))
lines(x = yrs, y = ints[2,], lty = 2) 
polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
text(1990, 3500, bquote(.(round(fixef(release_fit)[2] * 1000, 1))~lbs~year^-1))
@

\caption{There is an estimated increase in yield over time, but it is
  smaller than the previously estimated 54 lbs
  year\textsuperscript{-1}}
\label{fig:release}
\end{figure}

@@

@@

\begin{figure}
<<fig=TRUE, fig.width = 8, fig.height = 5, out.width = '80%', echo=FALSE>>=
tars <- c("M202","M206","M209")
par(mfrow = c(1,3),
    mar = c(4,0,2,0),
    oma = c(4,4,0.5,0.5))

invisible(sapply(tars, function(x){
    plot(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
         xlim = c(1995, 2016), ylim = c(6000,12000),
         pch = 16, main = x,
         ylab = "",
         yaxt = "n",
         xlab = "Year")
    if(x == "M202"){
        axis(2)
        mtext(expression("Yield ( lbs acre"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
    }
    abline(lm(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)

@
\caption{The performance of three major cultivars at the RES site over
time.}
\label{fig:res_yrs}
\end{figure}

<<fig=TRUE, echo=FALSE>>=
par(mfrow = c(1,3),
    mar = c(4,0,2,0),
    oma = c(4,4,0.5,0.5))

invisible(sapply(tars, function(x){
    plot(yield_lb ~ yrs_in_trial,
         data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
         xlim = c(-6, 36), ylim = c(6000,12000),
         pch = 16, main = x,
         ylab = "",
         yaxt = "n",
         xlab = "Years after release")
    if(x == "M202"){
        axis(2)
        mtext(expression("Yield ( lbs acre"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
    }
    abline(lm(yield_lb ~ yrs_in_trial,
              data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
}))

@
\caption{The performance of three major cultivars at the RES site
  since the year each cultivar was released.}
\label{fig:res_since_release}
\end{figure}

>>>>

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage[font=it]{caption}
\usepackage{Sweave}
\linespread{1.3}

@@

\section{Abstract}

Using the variety trial data, we find consistent estimates of the
reduction of yield once a variety is released, estimated to be between
-12 to -22 kg per year. This effect persists using all data, only
named/released cultivars, or only released cultivars only at the Rice
experiment station. This effect is offset by yield improvements over
the years, estimated to be small but positive effect of

\section{Introduction}

Constraints on ariable land are increasing simultaneous with the need to
increase total production to meet growing demand, which has necessitated
harvesting more grain per unit land area. Historically,
agricultural research has been successful in staving off the Mathesian
disaster through continued yield improvement of staple crops, however more and more
production systems are experiencing
plateaus in grain yield \citep{cassman}. If continued improvements per
unit land area are no longer possible, increasing the area under
cultivation will be needed to meet demand, which carries with it
undesirable ecological implications. Therefore, it is of critical
importance to better understand why grain yields have not continued to
increase.

CA rice production represents one such rice production
system. Rice yields in CA during the period from 1920 - 1990 increased rapidly,
but since 1990 have increased only slowly if at all (Fig. 1
\ref{fig:avg_yields}). However, continued improvement of grain yield as the
result of plant breeding has been demonstrated since 1990
\citep{rrb}, with each new cultivar havings a yield advantage over the
previous cultivars. We sought to better understand how grain yields
have increased in the breeders' fields without a corresponding
increase in yield in the farmers' fields. Two competing hypotheses
were tested: (1)
In the course of
investigating the drivers of yield in this system, we were compelled
to explore how yields could increase with each release of improved
cultivars, yet average yield for the region remain flat. Specifically,
we offered two competing hypotheses to explain this phenomenon: (1)
the yield advantage offered by newly developed cultivars is not
greater than previous cultivars, or (2) yield increases gained through
plant breeding are being offset by decreases in yield of released
cultivars.

Yet, annual reports from the CA
Cooperative Rice Research Foundation show that each released cultivar
has a yield advantage over the previous cultivars, typically on the
order of 4--6\%. This raises the question of how to reconcile these
facts. Namely, there are two possibilities:

@@

plot(yield_lb ~ yrs_in_trial, data = mod_data, pch = 16, col = rgb(0,0,0,0.25),
     xlab = "Year since entry",
     ylab = expression("Yield ( lbs acre"^-1~")"))
lines(x = yrs_in, y = ints[2,], lty = 2) 
polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
text(30, 4500, bquote(.(round(fixef(yrs_fit)[2] * 1000, 1))~lbs~year^-1))

@ 
\caption{There is a small decrease in yield over the time since a
  cultivar was entered into the trial using the full data
  set. However, the effect is small (\Sexpr{round(fixef(yrs_fit)[2] * 1000, 1)}
  lbs year\textsuperscript{-1}) in relation to the variability
  between sites and years.}
\label{fig:yrs}
\end{figure}

<<echo=FALSE>>=
all_CI <- round(posterior_interval(yrs_fit, pars = "yrs_in_trial") * 1000, 1)
RES_CI <- round(posterior_interval(RES_yrs_fit, pars = "yrs_in_trial") * 1000, 1)
@

As seen in Fig. \ref{fig:yrs}, there is a small decrease in yield
over time once a cultivar is released, though the estimated effect is
small compared to the variability in the data set. This decrease is estimated to be
\Sexpr{round(fixef(yrs_fit)[2] * 1000, 1)} lbs
year\textsuperscript{-1}. The oldest cultivar in the trials has lost
an estimated \Sexpr{round(fixef(yrs_fit)[2] * 1000 * 36, 1)} lbs
acre\textsuperscript{-1} since release. Using just data from the Rice
Experiment Station (RES), the estimated effect is more drastic
(an estimated \Sexpr{round(fixef(RES_yrs_fit)[2] * 1000, 1)} lbs
year\textsuperscript{-1}), but with a wider credible interval
(\Sexpr{RES_CI[1]} -- \Sexpr{RES_CI[2]} vs \Sexpr{all_CI[1]} --
\Sexpr{all_CI[2]}).

<<fig=TRUE, echo=FALSE>>=
yrs_in <- -6:34
pp <- posterior_predict(RES_yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
pp <-  pp * 1000 + 9000

ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))
RES <- mod_data[,'site'] == "RES"

plot(yield_lb ~ yrs_in_trial, data = mod_data[RES,], pch = 16, col = rgb(0,0,0,0.25),
     xlab = "Year since entry",
     ylab = expression("Yield ( lbs acre"^-1~")"))
lines(x = yrs_in, y = ints[2,], lty = 2) 
polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
text(30, 5000, bquote(.(round(fixef(RES_yrs_fit)[2] * 1000, 1))~lbs~year^-1))

@
\caption{The relationship between yield and years since entry for only
a single site, the Rice Experiment Station (RES). RES is considered
the best managed site, and has the most plots per year of any of the
sites. The estimated decrease over time is estimated to be stronger
using only these data (\Sexpr{round(fixef(RES_yrs_fit)[2] * 1000, 1)}
lbs year\textsuperscript{-1}), primarly due to lower variability.}
\label{fig:RES_yrs}
\end{figure}

\begin{figure}
<<fig=TRUE, fig.width = 8, fig.height = 5, out.width = '80%', echo=FALSE>>=
tars <- c("M202","M206")
par(mfrow = c(1, length(tars)),
    mar = c(4,0,2,0),
    oma = c(4,4,0.5,0.5))

invisible(sapply(tars, function(x){
    plot(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
         xlim = c(1995, 2016), ylim = c(5000,14000),
         pch = 16, main = x,
         ylab = "",
         yaxt = "n",
         xlab = "Year")
    if(x == "M202"){
        axis(2)
        mtext(expression("Yield ( lbs acre"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
    }
    abline(lm(yield_lb ~ year, data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)

@
\caption{The performance of two major cultivars at the RES site over
time. There is evidence that a cultivar's performance decreases over
time.}
\label{fig:res_yrs}
\end{figure}

<<fig=TRUE, echo=FALSE>>=
par(mfrow = c(1, length(tars)),
    mar = c(4,0,2,0),
    oma = c(4,4,0.5,0.5))

invisible(sapply(tars, function(x){
    plot(yield_lb ~ yrs_in_trial,
         data = mod_data[(mod_data$id == x & mod_data$site == "RES"),],
         xlim = c(-6, 36), ylim = c(5000,14000),
         pch = 16, main = x,
         ylab = "",
         yaxt = "n",
         xlab = "Years after release")
    if(x == "M202"){
        axis(2)
        mtext(expression("Yield ( lbs acre"^-1~")"), 2, line = 2.25, cex.lab = 0.6)
    }
    abline(lm(yield_lb ~ yrs_in_trial,
              data = mod_data[(mod_data$id == x & mod_data$site == "RES"),]), lty = 2)
}))

@
\caption{The performance of two major cultivars at the RES site
  since the year each cultivar was released. Both M-202 and M-206 have
  similar intercepts, suggesting similar yield at the time of release.}
\label{fig:res_since_release}
\end{figure}

@@

plot(yield_lb ~ release_yr, data = mod_data, pch = 16, col = rgb(0,0,0,0.25),
     xlab = "Year of Release",
     ylab =  expression("Yield ( lbs acre"^-1~")"))
lines(x = yrs, y = ints[2,], lty = 2) 
polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
text(2010, 4500, bquote(.(round(fixef(release_fit)[2] * 1000, 1))~lbs~year^-1))
@

\caption{The effect of year of release on yield. There is an estimated
  increase in yield over time, but it is smaller than the previously estimated 54 lbs
  year\textsuperscript{-1}}
\label{fig:release}
\end{figure}

@@

<<fig=TRUE, echo=FALSE>>=
yrs <- 1980:2015
pp <- posterior_predict(RES_rel_fit, re.form = NA,
                        newdata = data.frame(release_yr_c = yrs - 2000))
pp <-  pp * 1000 + 9000

plot(yield_lb ~ release_yr, data = mod_data[RES,], pch = 16, col = rgb(0,0,0,0.25),
     xlab = "Year of Release",
     ylab =  expression("Yield ( lbs acre"^-1~")"))
lines(x = yrs, y = ints[2,], lty = 2) 
polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
text(2010, 4500, bquote(.(round(fixef(RES_rel_fit)[2] * 1000, 1))~lbs~year^-1))
@

\caption{The effect of year of release on yield for only data from the
  Rice Experiment Station (RES). There is an estimated
  increase in yield over time, but it is smaller than the previously estimated 54 lbs
  year\textsuperscript{-1}}
\label{fig:RES_release}
\end{figure}

@@
